Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2018-02-19 revised_prop_letter_ajhs.tex
From: Tai, Yu-Tsung Sent: Monday, February 19, 2018 5:29 To: Hanson, Andrew J. <hansona@indiana.edu>; Ortiz, Gerardo <ortizg@indiana.edu> Cc: Sabry, Amr A. <sabry@indiana.edu> Subject: RE: letter to the editor
Dear all,

I changed the text according to my previous email, and the source files are in the ShareLaTeX and GitHub as usual. Please check whether it looks fine and see if you have any suggestions… Thanks : )

Best regards,
Tai, Yu-Tsung (戴淯琮)
PhD Candidate double-majoring in Mathematics & Computer Science
Indiana University, Bloomington
https://www.linkedin.com/in/yu-tsung-tai-9aa30551

Let $E_1$ and $E_2$ be two disjoint events with probabilities
$\mu(E_1)=[\ell_1,r_1]$ and $\mu(E_2)=[\ell_2,r_2]$. A first attempt
at calculating the probability of the combined event that
\emph{either} $E_1$ or $E_2$ occurs might be
$\mu(E_1\cup E_2) = [\ell_1+\ell_2,r_1+r_2]$. In some cases, this is
indeed a sensible definition. For example, if $\mu(E_1)=[0.1,0.2]$ and
$\mu(E_2)=[0.3,0.4]$ we get $\mu(E_1\cup E_2) = [0.4,0.6]$. But
consider an event $E$ such that $\mu(E)=[0.2,0.3]$ and hence
$\mu\left(\overline{E}\right)=[0.7,0.8]$. The two events $E$ and $\overline{E}$
are disjoint; the naïve addition of intervals would give
$\mu\left(E\cup\overline{E}\right)=[0.9,1.1]$ which is not a valid probability
interval. Moreover the event $E\cup\overline{E}$ is the
entire space; its probability interval should be
$\necess$ which is sharper than $[0.9,1.1]$. The problem is that the
two intervals are correlated: there is more information in the
combined event than in each event separately. In our example, even
though the ``true'' probability of $E$ can be anywhere in the range
$[0.2,0.3]$ and the ``true'' probability of~$\overline{E}$ can be
anywhere in the range $[0.7,0.8]$, the values are not independent. Any
value of $\mu(E) \leq 0.25$ will force $\mu\left(\overline{E}\right)\geq
0.75$. To account for such subtleties, the axioms of interval-valued
probability do not use a strict equality for the combination of
disjoint events. The correct constraint enforcing coherence of
the probability assignment for
$E_1\cup E_2$ when $E_1$ and $E_2$ are disjoint is therefore:
\begin{equation}
\label{eq:disjoint}
\mu(E_1\cup E_2) \subseteq [\ell_1+\ell_2,r_1+r_2]\,.
\end{equation}
Note that for any event $E$ with $\mu(E)=[\ell,r]$, we always have
$\mu(\Omega)=\necess\subseteq[\ell,r]+[1-r,1-\ell]=\mu(E)\cup\mu\left(\overline{E}\right)$.
%% Parallel arguments
%% can be made interpreting $[\ell,r]$ as Gaussian width, but will be omitted as no
%% essential new information seems to be introduced.

@@

This is Meyer's great insight and the main assumption in his proof that using the
field of rationals nullifies the Kochen-Specker theorem.  In other
words, Meyer exclusively attributes finite-precision errors to the
description of the states defined in a general Hilbert space. The
response of Mermin, argues against Meyer's interpretation but accepts
his framework. More generally, there have been responses that argue
that the effect of finite-precision measurements on the Kochen-Specker
theorem requires a different approach altogether (see for example Ax
and Kochen's communication cited by Cabello~\cite{Cabello2002}). Our
approach is agnostic about attributing elements of reality to the
state function (as Meyer did). We simply attribute lack of certainty
to the uncertain results obtained through the use of a measuring
instrument. Thus, the parameter~$\delta$ in our framework reflects
insufficient knowledge of the experimenter, which could be due to a
variety of reasons related to imperfections of devices. For example, a
typical question that an experimenter may not be able to answer
accurately would be ``did an electron land in the left half or the
right half of the screen?''  There are cases in which the electron
would land too close to the middle of the screen for the experimenter
to be able to determine with certainty that it was left or right of
the center. We simply record this imprecision in the probability we
assign to events, and use the axioms of probability theory to argue
that a sharp transition occurs when the imprecision reaches a certain
value.

@@

To see this idea in the context of a quantum experiment, consider
a three-dimensional Hilbert space with
one-dimensional projectors~$P_\rho$, two-dimensional projectors
$P_\rho + P_\sigma$, and an experiment that is experiment repeated $12$
times. By the Kochen-Specker theorem, it is impossible to build a probability
measure that maps every projection to either $0=\frac{0}{12}$ or
$1=\frac{12}{12}$. That is, the following assignment~$\bar{\mu}_{0}$
is not a QIVPM:
\begin{center}
\begin{tabular}{cc}
\toprule 
\addlinespace
$P$  & $\bar{\mu}_{0}\left(P\right)$\tabularnewline
\midrule
\midrule 
\addlinespace
$\mathbb{0}$  & $\imposs$\tabularnewline
\midrule 
\addlinespace
All one-dimensional projectors  & $\left[0,0\right]$\tabularnewline
\midrule 
\addlinespace
All two-dimensional projectors  & $\left[1,1\right]$\tabularnewline
\midrule 
\addlinespace
$\mathbb{1}$  & $\necess$\tabularnewline
\bottomrule
\end{tabular}
\par\end{center}

Now consider what happens if $\frac{1}{4}$ of the data for \emph{every}
one-dimensional projector is uncertain. A potential account of this
degradation is to assign to each event $P$ the entire range of possibilities
$\bar{\mu}_{1}(P)$: 
\begin{center}
\begin{tabular}{cc}
\toprule 
\addlinespace
$P$  & $\bar{\mu}_{1}\left(P\right)$\tabularnewline
\midrule
\midrule 
\addlinespace
$\mathbb{0}$  & $\imposs$\tabularnewline
\midrule 
\addlinespace
All one-dimensional projectors  & $\left[0,\tfrac{1}{4}\right]$\tabularnewline
\midrule 
\addlinespace
All two-dimensional projectors  & $\left[\tfrac{3}{4},1\right]$\tabularnewline
\midrule 
\addlinespace
$\mathbb{1}$  & $\necess$\tabularnewline
\bottomrule
\end{tabular}
\par\end{center}

\noindent This measure is not a valid QIVPM because it does not satisfy
the convexity condition: for any two orthogonal one-dimensional events
$P_{0}$ and $P_{1}$, the convexity condition requires $\bar{\mu}_{1}\left(P_{0}+P_{1}\right)\subseteq\bar{\mu}_{1}\left(P_{0}\right)+\bar{\mu}_{1}\left(P_{1}\right)$,
but $\bar{\mu}_{1}\left(P_{0}+P_{1}\right)=\left[\tfrac{3}{4},1\right]$
which is not a subset of $\left[0,\tfrac{1}{2}\right]=\bar{\mu}_{1}\left(P_{0}\right)+\bar{\mu}_{1}\left(P_{1}\right)$.
Interestingly, it is impossible to find any probability measure that
would be consistent with these observations, as the interval $\left[\tfrac{3}{4},1\right]$
is completely disjoint from the interval $\left[0,\tfrac{1}{2}\right]$
and no amount of shifting of assumptions regarding the precise outcome
of the uncertain observations could change that disjointness. However,
as shown next, a sharp transition occurs when $\delta=\tfrac{1}{3}$.

When the proportion of uncertain data reaches $\frac{1}{3}$, the probability measure
that assigns to each event the entire range of possibilities is: 
\begin{center}
\begin{tabular}{cc}
\toprule 
\addlinespace
$P$  & $\bar{\mu}_{2}\left(P\right)$\tabularnewline
\midrule
\midrule 
\addlinespace
$\mathbb{0}$  & $\imposs$\tabularnewline
\midrule 
\addlinespace
All one-dimensional projectors  & $\left[0,\tfrac{1}{3}\right]$\tabularnewline
\midrule 
\addlinespace
All two-dimensional projectors  & $\left[\tfrac{2}{3},1\right]$\tabularnewline
\midrule 
\addlinespace
$\mathbb{1}$  & $\necess$\tabularnewline
\bottomrule
\end{tabular}
\par\end{center}

\noindent This is also not a valid probability measure by the same
argument as above. However, in this case $\bar{\mu}_{2}\left(P_{0}+P_{1}\right)=\left[\tfrac{2}{3},1\right]$
and $\left[0,\tfrac{2}{3}\right]=\bar{\mu}_{2}\left(P_{0}\right)+\bar{\mu}_{2}\left(P_{1}\right)$
have a \emph{common point}. Hence, by assuming that the uncertain data for
one-dimensional projectors always support the associated event, while
those for two-dimensional projectors always refute the event, we can
find the following probability measure that can verified to be a
valid QIVPM and is consistent with the experimental data: 
\begin{center}
\begin{tabular}{cc}
\toprule 
\addlinespace
$P$  & $\bar{\mu}_{2}'\left(P\right)$\tabularnewline
\midrule
\midrule 
\addlinespace
$\mathbb{0}$  & $\imposs$\tabularnewline
\midrule 
\addlinespace
All one-dimensional projectors  & $\left[\tfrac{1}{3},\tfrac{1}{3}\right]$\tabularnewline
\midrule 
\addlinespace
All two-dimensional projectors  & $\left[\tfrac{2}{3},\tfrac{2}{3}\right]$\tabularnewline
\midrule 
\addlinespace
$\mathbb{1}$  & $\necess$\tabularnewline
\bottomrule
\end{tabular}
\par\end{center}

A similar situation happens when more than $\frac{1}{3}$ of data
is uncertain. In particular, if half of the data is uncertain, the probability
measure that assigns to each event the entire range of possibilities
is already a QIVPM:
\begin{center}
\begin{tabular}{cc}
\toprule 
\addlinespace
$P$  & $\bar{\mu}_{3}\left(P\right)$\tabularnewline
\midrule
\midrule 
\addlinespace
$\mathbb{0}$  & $\imposs$\tabularnewline
\midrule 
\addlinespace
All one-dimensional projectors  & $\left[0,\tfrac{1}{2}\right]$\tabularnewline
\midrule 
\addlinespace
All two-dimensional projectors  & $\left[\tfrac{1}{2},1\right]$\tabularnewline
\midrule 
\addlinespace
$\mathbb{1}$  & $\necess$\tabularnewline
\bottomrule
\end{tabular}
\par\end{center}

>>>>

Let $E_1$ and $E_2$ be two disjoint events with probabilities
$\mu(E_1)=[\ell_1,r_1]$ and $\mu(E_2)=[\ell_2,r_2]$. A first attempt
at calculating the probability of the combined event that
\emph{either} $E_1$ or $E_2$ occurs might be
$\mu(E_1\cup E_2) = [\ell_1+\ell_2,r_1+r_2]$. In some cases, this is
indeed a sensible definition. For example, if $\mu(E_1)=[0.1,0.2]$ and
$\mu(E_2)=[0.3,0.4]$ we get $\mu(E_1\cup E_2) = [0.4,0.6]$. But
consider an event $E$ such that $\mu(E)=[0.2,0.3]$ and hence
$\mu\left(\overline{E}\right)=[0.7,0.8]$. The two events $E$ and $\overline{E}$
are disjoint; the naïve addition of intervals would give
$\mu\left(E\cup\overline{E}\right)=[0.9,1.1]$ which is not a valid probability
interval. Moreover the event $E\cup\overline{E}$ is the
entire space; its probability interval should be
$\necess$ which is sharper than $[0.9,1.1]$. The problem is that the
two intervals are correlated: there is more information in the
combined event than in each event separately so the combined event should be mapped to
a sharper interval. In our example, even
though the ``true'' probability of $E$ can be anywhere in the range
$[0.2,0.3]$ and the ``true'' probability of~$\overline{E}$ can be
anywhere in the range $[0.7,0.8]$, the values are not independent. Any
value of $\mu(E) \leq 0.25$ will force $\mu\left(\overline{E}\right)\geq
0.75$. To account for such subtleties, the axioms of interval-valued
probability do not use a strict equality for the combination of
disjoint events. The correct constraint enforcing coherence of
the probability assignment for
$E_1\cup E_2$ when $E_1$ and $E_2$ are disjoint is therefore:
\begin{equation}
\label{eq:disjoint}
\mu(E_1\cup E_2) \subseteq [\ell_1+\ell_2,r_1+r_2]\,.
\end{equation}
Note that for any event $E$ with $\mu(E)=[\ell,r]$, we always have
$\mu(\Omega)=\necess\subseteq[\ell,r]+[1-r,1-\ell]=\mu(E)\cup\mu\left(\overline{E}\right)$.
%% Parallel arguments
%% can be made interpreting $[\ell,r]$ as Gaussian width, but will be omitted as no
%% essential new information seems to be introduced.

@@

\begin{table*}
\caption{\label{tab:probability-measures}Possible probability measures on
a Hilbert space of dimension~$d=3$, where $\bar{\mu}_{2}'$ and
$\bar{\mu}_{3}$ are QIVPMs while $\bar{\mu}_{0}$, $\bar{\mu}_{1}$,
and $\bar{\mu}_{2}$ are not. Events are listed in the column labeled
by $P$.}
\centering{}%
\begin{tabular}{cccccc}
\toprule 
\addlinespace
$P$  & $\bar{\mu}_{0}\left(P\right)$ & $\bar{\mu}_{1}\left(P\right)$ & $\bar{\mu}_{2}\left(P\right)$ & $\bar{\mu}_{2}'\left(P\right)$ & $\bar{\mu}_{3}\left(P\right)$\tabularnewline
\midrule
\midrule 
\addlinespace
$\mathbb{0}$  & $\imposs$ & $\imposs$ & $\imposs$ & $\imposs$ & $\imposs$\tabularnewline
\midrule 
\addlinespace
All one-dimensional projectors  & $\left[0,0\right]$ & $\left[0,\tfrac{1}{4}\right]$ & $\left[0,\tfrac{1}{3}\right]$ & $\left[\tfrac{1}{3},\tfrac{1}{3}\right]$ & $\left[0,\tfrac{1}{2}\right]$\tabularnewline
\midrule 
\addlinespace
All two-dimensional projectors  & $\left[1,1\right]$ & $\left[\tfrac{3}{4},1\right]$ & $\left[\tfrac{2}{3},1\right]$ & $\left[\tfrac{2}{3},\tfrac{2}{3}\right]$ & $\left[\tfrac{1}{2},1\right]$\tabularnewline
\midrule 
\addlinespace
$\mathbb{1}$  & $\necess$ & $\necess$ & $\necess$ & $\necess$ & $\necess$\tabularnewline
\bottomrule
\end{tabular}
\end{table*}
This is Meyer's great insight and the main assumption in his proof that using the
field of rationals nullifies the Kochen-Specker theorem.  In other
words, Meyer exclusively attributes finite-precision errors to the
description of the states defined in a general Hilbert space. The
response of Mermin, argues against Meyer's interpretation but accepts
his framework. More generally, there have been responses that argue
that the effect of finite-precision measurements on the Kochen-Specker
theorem requires a different approach altogether (see for example Ax
and Kochen's communication cited by Cabello~\cite{Cabello2002}). Our
approach is agnostic about attributing elements of reality to the
state function (as Meyer did). We simply attribute lack of certainty
to the uncertain results obtained through the use of a measuring
instrument. Thus, the parameter~$\delta$ in our framework reflects
insufficient knowledge of the experimenter, which could be due to a
variety of reasons related to imperfections of devices. For example, a
typical question that an experimenter may not be able to answer
accurately would be ``did an electron land in the left half or the
right half of the screen?''  There are cases in which the electron
would land too close to the middle of the screen for the experimenter
to be able to determine with certainty that it was left or right of
the center. We simply record this imprecision in the probability we
assign to events, and use the axioms of probability theory to argue
that a sharp transition occurs when the imprecision reaches a certain
value.

@@

To see this idea in the context of a quantum experiment, consider
a three-dimensional Hilbert space with
one-dimensional projectors~$P_{\rho}$, two-dimensional projectors
$P_{\rho}+P_{\sigma}$, and an experiment that is experiment repeated $12$
times. By the Kochen-Specker theorem, it is impossible to build a probability
measure that maps every projection to either $0=\frac{0}{12}$ or
$1=\frac{12}{12}$. That is, the assignment~$\bar{\mu}_{0}$
defined in Table~\ref{tab:probability-measures} is not a QIVPM.

Now consider what happens if $\frac{1}{4}$ of the data for \emph{every}
one-dimensional projector is uncertain. A potential account of this
degradation is to assign to each event $P$ the entire range of possibilities
$\bar{\mu}_{1}(P)$ as defined in Table~\ref{tab:probability-measures}.
This measure is not a valid QIVPM because it does not satisfy
the convexity condition: for any two orthogonal one-dimensional events
$P_{0}$ and $P_{1}$, the convexity condition requires $\bar{\mu}_{1}\left(P_{0}+P_{1}\right)\subseteq\bar{\mu}_{1}\left(P_{0}\right)+\bar{\mu}_{1}\left(P_{1}\right)$,
but $\bar{\mu}_{1}\left(P_{0}+P_{1}\right)=\left[\tfrac{3}{4},1\right]$
which is not a subset of $\left[0,\tfrac{1}{2}\right]=\bar{\mu}_{1}\left(P_{0}\right)+\bar{\mu}_{1}\left(P_{1}\right)$.
Interestingly, it is impossible to find any probability measure that
would be consistent with these observations, as the interval $\left[\tfrac{3}{4},1\right]$
is completely disjoint from the interval $\left[0,\tfrac{1}{2}\right]$
and no amount of shifting of assumptions regarding the precise outcome
of the uncertain observations could change that disjointness. However,
as shown next, a sharp transition occurs when $\delta=\tfrac{1}{3}$.

When the proportion of uncertain data reaches $\frac{1}{3}$, the probability measure
that assigns to each event the entire range of possibilities is
$\bar{\mu}_{2}$ defined in Table~\ref{tab:probability-measures}.
This is also not a valid probability measure by the same
argument as above. However, in this case $\bar{\mu}_{2}\left(P_{0}+P_{1}\right)=\left[\tfrac{2}{3},1\right]$
and $\left[0,\tfrac{2}{3}\right]=\bar{\mu}_{2}\left(P_{0}\right)+\bar{\mu}_{2}\left(P_{1}\right)$
have a \emph{common point}. Hence, by assuming that the uncertain data for
one-dimensional projectors always support the associated event, while
those for two-dimensional projectors always refute the event, we can
find the probability measure~$\bar{\mu}_{2}'$ that can verified to be a
valid QIVPM and is consistent with the experimental data.

A similar situation happens when more than $\frac{1}{3}$ of data
is uncertain. In particular, if half of the data is uncertain, the probability
measure~$\bar{\mu}_{3}$ that assigns to each event the entire range of possibilities
is already a QIVPM.
